External Validation

### 2021-02-17 FEM_R/model_analysis.Rmd
Added lots more validation to RNotebook file, and added some of these to model_analysis.Rmd

@@

>>>>

Second go at this cross-validation, attempting to solve some issues by using the 60 plus population.

  # Now lets extract chronic disease variables from both dataframes and do a row combine
  # Start with cancre, diabe, hearte
  cd_l2 <- select(l2, c(year, age, cancre, diabe, hearte, lunge, smoken, bmi, employed, anyadl)) #, anyiadl, adl1, adl2, adl3p))
  cd_CV2 <- select(CV2, c(year, a_age_60p, p_cancre_60p, p_diabe_60p, 
                            p_hearte_60p, p_lunge_60p, p_smoken_60p, 
                            a_bmi_60p, p_employed_60p,  p_anyadl_60p))
                            #, p_anyiadl_60p p_adl1_60p, p_adl2_60p, p_adl3p_60p))

  CV2_rename <- cd_CV2 %>% dplyr::rename('age' = a_age_60p,
                                        'cancre' = p_cancre_60p, 
                                        'diabe' = p_diabe_60p, 
                                        'hearte' = p_hearte_60p,
                                        'lunge' = p_lunge_60p,
                                        'smoken' = p_smoken_60p,
                                        'bmi' = a_bmi_60p,
                                        'employed' = p_employed_60p,
                                        'anyadl' = p_anyadl_60p)
                                        #'anyiadl' = p_anyiadl_60p)
                                        #'adl1' = p_adl1_60p,
                                        #'adl2' = p_adl2_60p,
                                        #'adl3p' = p_adl3p_60p)

@@

# Smoking

```{r}
FEM_base <- read_dta('ELSA_Baseline/ELSA_Baseline_summary.dta')

# Smoking prevalence stats
smok.5059 <- 15.2
smok.60p <- 10.2

# FEM summary, filter for 2018 and 60p
#FEM.smok <- FEM_sum
FEM.smok <- FEM_base
FEM.smok <- FEM.smok %>%
                filter(year == 2018) %>%
                select(contains('smoken')) %>%
                select(contains('60p'), contains('5059')) %>%
                select(contains('p_'))

FEM.smok <- FEM.smok * 100

smok.combined <- data.frame(FEM = FEM.smok[1,1], AgeUK = 10.2)
smok.combined <- data.frame(source = c('FEM', 'FEM', 'Observed', 'Observed'), age = c('50 - 59', '60+', '50 - 59', '60+'), value = c(FEM.smok[1,2], FEM.smok[1,1], 15.2, 10.2))

smok.plot <-ggplot(data = smok.combined, aes(x = age, y = value)) +
                geom_bar(aes(fill = source), stat = "identity", position = "dodge") +
                labs(title = 'Comparison of smoking prevalence in the elderly in 2018', x = ' ', y = 'Prevalence (%)')

ggsave(filename = 'smoking_prev_2018.png',
       plot = smok.plot,
       path = '/home/luke/Documents/E_FEM_clean/Deliverables/E-FEM_Paper/figures/Validation/',
       width = 7,
       height = 5)

smok.plot

### 2021-02-17 FEM_R/validation.Rmd
Added lots more validation to RNotebook file, and added some of these to model_analysis.Rmd

require(tidyverse)
#require(reactable)
require(haven)
#require(data.table)
require(reshape2)
require(ggplot2)
#require(plyr)

@@

# FEM 2014 summary
FEM_sum <- read_dta('ELSA_Baseline/ELSA_Baseline_summary.dta')

@@

  # Save plot into in E-FEM_Paper/figures/
  ggsave(filename = paste0(cd, '_', gender, '.png'),
         plot = p,
         path = path,
         width = 7,
         height = 5)

@@

>>>>

require(tidyverse)
#require(reactable)
require(haven)
#require(data.table)
require(reshape2)
require(ggplot2)
#require(plyr)
require(readODS)

@@

# FEM 2014 summary
FEM_sum <- read_dta('ELSA_AgeUK_valid/ELSA_AgeUK_valid_summary.dta')

@@

  # Save plot into in E-FEM_Paper/figures/
  #ggsave(filename = paste0(cd, '_', gender, '.png'),
  #       plot = p,
  #       path = path,
  #       width = 7,
  #       height = 5)

@@

# Smoking

## Action on Smoking and Health

```{r}
FEM_base <- read_dta('ELSA_Baseline/ELSA_Baseline_summary.dta')

# Smoking prevalence stats
# Stats found here: https://ash.org.uk/information-and-resources/fact-sheets/smoking-statistics-who-smokes-and-how-much/
# Page 3, table 4
smok.5059 <- 15.2
smok.60p <- 10.2

# FEM summary, filter for 2018 and 60p
#FEM.smok <- FEM_sum
FEM.smok <- FEM_base
FEM.smok <- FEM.smok %>%
                filter(year == 2018) %>%
                select(contains('smoken')) %>%
                select(contains('60p'), contains('5059')) %>%
                select(contains('p_'))

FEM.smok <- FEM.smok * 100

smok.combined <- data.frame(FEM = FEM.smok[1,1], AgeUK = 10.2)
smok.combined <- data.frame(source = c('FEM', 'FEM', 'Observed', 'Observed'), age = c('50 - 59', '60+', '50 - 59', '60+'), value = c(FEM.smok[1,2], FEM.smok[1,1], 15.2, 10.2))

smok.plot <-ggplot(data = smok.combined, aes(x = age, y = value)) +
                geom_bar(aes(fill = source), stat = "identity", position = "dodge") +
                labs(title = 'Comparison of smoking prevalence in the elderly in 2018', x = ' ', y = 'Prevalence (%)')

ggsave(filename = 'smoking_prev_2018.png',
       plot = smok.plot,
       path = '/home/luke/Documents/E_FEM_clean/Deliverables/E-FEM_Paper/figures/Validation/',
       width = 7,
       height = 5)

smok.plot

## ONS

Found some more data on smoking in the UK from the ONS, perhaps more reliable than the other? Might be a good idea to show both if they are all within a couple of percent.

```{r}
# FEM
FEM_base <- read_dta('ELSA_Baseline/ELSA_Baseline_summary.dta')

# ONS - Male
m.ons.smok <- read_ods(path = '/home/luke/Documents/E_FEM_clean/Validation/adultsmokinghabitsinengland2019final.ods',
                sheet = 4,
                skip = 8,
                col_names = TRUE,
                row_names = FALSE,
                range = "E1:F22")
# ONS - Female
f.ons.smok <- read_ods(path = '/home/luke/Documents/E_FEM_clean/Validation/adultsmokinghabitsinengland2019final.ods',
                sheet = 4,
                skip = 8,
                col_names = TRUE,
                row_names = FALSE,
                range = "L1:M22")
# ONS - All
all.ons.smok <- read_ods(path = '/home/luke/Documents/E_FEM_clean/Validation/adultsmokinghabitsinengland2019final.ods',
                sheet = 4,
                skip = 8,
                col_names = TRUE,
                row_names = FALSE,
                range = "S1:T22")

# Remove empty row
m.ons.smok <- m.ons.smok %>% slice(-1)
f.ons.smok <- f.ons.smok %>% slice(-1)
all.ons.smok <- all.ons.smok %>% slice(-1)

# Add year column
m.ons.smok$Year <- c(2000:2019)
f.ons.smok$Year <- c(2000:2019)
all.ons.smok$Year <- c(2000:2019)

# Add Gender col
m.ons.smok$Gender <- 'Male'
f.ons.smok$Gender <- 'Female'
all.ons.smok$Gender <- 'All'

# Rearrange columns before rbind
m.ons.smok <- subset(m.ons.smok, select = c(3,4,1,2))
f.ons.smok <- subset(f.ons.smok, select = c(3,4,1,2))
all.ons.smok <- subset(all.ons.smok, select = c(3,4,1,2))

# Filter out years before 2012
m.ons.smok <- dplyr::filter(m.ons.smok, Year >= 2012)
f.ons.smok <- dplyr::filter(f.ons.smok, Year >= 2012)
all.ons.smok <- dplyr::filter(all.ons.smok, Year >= 2012)

# Combine male, female, all and rename 60 plus column
ons.smok <- rbind(m.ons.smok, f.ons.smok, all.ons.smok)
ons.smok <- dplyr::rename(ons.smok, '60p' = '60 and over')

# Reshape data to long
pivot.ons.smok <- pivot_longer(data = ons.smok,
                               cols = '50-59':'60p',
                               names_to = 'Age',
                               values_to = 'Value')

pivot.ons.smok$Source <- 'ONS'

# ONS data wrangled, now get FEM into shape
FEM.smok <- FEM_base
# Get year out to put back later
Year <- FEM.smok$year
FEM.smok <- FEM.smok %>%
                select(contains('smoken')) %>%
                select(contains('p_'), -starts_with('n_')) %>%
                select(contains('60p'), contains('5059')) %>%
                dplyr::rename('60p' = p_smoken_60p, '50-59' = p_smoken_5059)

FEM.smok <- subset(FEM.smok, select = c(2, 1))
FEM.smok <- cbind(Year, FEM.smok)

FEM.smok <- filter(FEM.smok, year >= 2012, year < 2022)

pivot.FEM.smok <- pivot_longer(data = FEM.smok,
                               cols = '50-59':'60p',
                               names_to = 'Age',
                               values_to = 'Value')
pivot.FEM.smok$Source <- 'FEM'
pivot.FEM.smok$Value <- pivot.FEM.smok$Value * 100

pivot.ons.smok.all <- pivot.ons.smok %>%
                          filter(Gender == 'All') %>%
                          select(-Gender)

combined <- rbind(pivot.FEM.smok, pivot.ons.smok.all)

ggplot(data = combined, aes(x = Year, y = Value)) +
  geom_bar(aes(fill = Source), stat = 'identity', position = 'dodge')

# ADLs

Just having a play around with this data, wanted to visualise how levels of anyADL change over time.

FEM.new <- FEM_base

FEM.adl <- FEM.new %>%
              select(contains('year'), contains('anyadl')) %>%
              select(-starts_with('i_'), -starts_with('n_'), -starts_with('a_')) %>%
              select(-contains('_m'), -contains('_f')) %>%
              select(-contains('5059'), -contains('60p'))

p.FEM.adl <- pivot_longer(data = FEM.adl,
                          cols = p_anyadl_100p:p_anyadl_all,
                          names_to = c('Prevalence', 'Condition', 'Age'),
                          names_sep = '_')

p.FEM.adl <- select(p.FEM.adl, -Prevalence, -Condition)

p.FEM.adl$Age <- factor(p.FEM.adl$Age, levels = c('6064', '6569', '7074', '7579', '8084', '8589', '9094', '9599', '100p', 'all'))

ggplot(data = p.FEM.adl, aes(x = year, y = value, colour = Age)) +
  geom_line(aes(fill = Age))

# ADL Comparison (Wittenberg)

Data for this comparison came from [this paper](https://people.uea.ac.uk/en/publications/projections-of-demand-and-expenditure-on-adult-social-care-2015-to-2040(d1f44389-4e52-4cb1-9b66-ecc7594d345e).html) - Projections of Demand and Expenditure on Adult Social Care 2015 to 2040. On page 6, base case projections are outlined. Wittenberg says that from 2015 to 2040, the population having at least 1 difficulty in ADL or IADL would rise by 67%, and from 2015 to 2070 would rise by 116%. When considering only ADLs, the increase was 69% from 2015 to 2040, and 124% from 2015 to 2070. Written more clearly:

ADL & IADL
2015 - 2040: 67% increase
2015 - 2070: 116% increase

ADL only
2015 - 2040: 69% increase
2015 - 2070: 124% increase

FEM_base <- read_dta('ELSA_Baseline/ELSA_Baseline_summary.dta')

# Get FEM output and extract only p_anyadl_all
FEM.adl <- FEM_base %>%
              select(contains('year'), contains('anyadl')) %>%
              select(-starts_with('i_'), -starts_with('n_'), -starts_with('a_')) %>%
              select(-contains('_m'), -contains('_f')) %>%
              select(contains('year'), contains('_all'))

# Now calculate % increase between 2015 & 2040. Need to take average between 2014 & 2016 values
FEM.2015 <- data.frame(year = 2015, ((FEM.adl[2,2] + FEM.adl[3,2]) / 2))
# extract 2040
FEM.2040 <- filter(FEM.adl, year == 2040)
# Row bind
FEM.2015_2040 <- rbind(FEM.2015, FEM.2040)

# Now calculate percentage change
FEM.adl.perchange <- ((FEM.2015_2040[2,2] - FEM.2015_2040[1,2]) / FEM.2015_2040[2,2]) * 100

# Not good! Showing an 8% reduction in anyadl when we should be seeing a ~70% increase
# Have changed some transition models to see if we can change this significantly
# Now 14% reduction, not good. Need to have a good look at these later.

# Stroke Comparison (Wittenberg)

Data from [this paper](https://www.semanticscholar.org/paper/The-future-incidence%2C-prevalence-and-costs-of-in-UK-King-Wittenberg/fdaffc8cd7790556e57c77c0b195c8f7a64c4189) by King et al. (2020). Their model predicts an increase in prevalence of 120% from 2015 to 2035. Also, paper projects that incidence of stroke will increase by 60% each year in the same period. Going to start by comparing the prevalence, but when I come back to this I think it would be interesting to compare incidence also.

FEM_base <- read_dta('ELSA_Baseline/ELSA_Baseline_summary.dta')

# Get FEM output and extract only p_anyadl_all
FEM.stroke <- FEM_base %>%
              select(contains('year'), contains('stroke')) %>%
              select(-starts_with('i_'), -starts_with('n_'), -starts_with('a_')) %>%
              select(contains('year'), contains('_all'))
# select(-contains('_m'), -contains('_f')) %>%

FEM.stroke.2015 <- data.frame(year = 2015, ((FEM.stroke[2,2] + FEM.stroke[3,2]) / 2))
FEM.stroke.2040 <- filter(FEM.stroke, year == 2040)
FEM.stroke.2015_2040 <- rbind(FEM.stroke.2015, FEM.stroke.2040)

# Now calculate percentage change
FEM.stroke.perchange <- ((FEM.stroke.2015_2040[2,2] - FEM.stroke.2015_2040[1,2]) / FEM.stroke.2015_2040[2,2]) * 100

# Output showing a ~29% increase in stroke, nowhere near the 120% increase Wittenberg predicts

FEM_base <- read_dta('ELSA_Baseline/ELSA_Baseline_summary.dta')

# Get FEM output and extract only p_anyadl_all
FEM.dem <- FEM_base %>%
              select(contains('year'), contains('demene')) %>%
              select(-starts_with('i_'), -starts_with('n_'), -starts_with('a_')) %>%
              select(contains('year'), contains('_all'))
# select(-contains('_m'), -contains('_f')) %>%

FEM.dem.2015 <- data.frame(year = 2015, ((FEM.dem[2,2] + FEM.dem[3,2]) / 2))
FEM.dem.2040 <- filter(FEM.dem, year == 2040)
FEM.dem.2015_2040 <- rbind(FEM.dem.2015, FEM.dem.2040)

# Now calculate percentage change
FEM.dem.perchange <- ((FEM.dem.2015_2040[2,2] - FEM.dem.2015_2040[1,2]) / FEM.dem.2015_2040[2,2]) * 100

# Bland Altman method of Validation

```{r}
# Contains method for bland-altman validation 
require(MethComp)
